---
title: "Data_Science_project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}

library(readr)
library(tidyverse)
library(ggplot2)
library(janitor)
library(tidycensus)
library(viridis) 
library(tscount)
library(yrbss)
library(zoo)
library(purrr)

```



## R Markdown

## Motivation and Objectives
In September, 2014, the Federal Bureau of Investigation (FBI) released a report “A Study of Active Shooter Incidents in the United States Between 2000 and 2013” stating that the incidence of school and mass shooting incidents has been growing over the past 14 years $^{[1, 2]}$. Very recently on November 14th, 2019, there was a shooting at a high school in Santa Clarita, California where a 16-year-old open fire at the school, killing 2 people and injuring 3 others $^{[3]}$. On average, mass killings involving firearms occur approximately every two weeks in the US, while school shootings occur on average monthly $^{[4]}$. According to an analysis by “Washington Post”, 233,000 students have experienced gun violence in school since the Columbine High massacre in 1999 and the number is increasing alarmingly$^{[5]}$. Such traumatic experiences leave a strong affect on the mental health of the survivors. Research indicates that exposure to assaultive violence, or learning that a close friend or loved one has faced such exposure, is associated with an increased incidence of a range of negative mental health outcomes, among them posttraumatic stress disorder (PTSD) and major depression $^{[6,7]}$. 


## Related Work
In order to effectively address gun violence in schools, it should be first acknowledged that the fact,it is a gun violence problem. Though there have been many comprehensive school safety plans proposed over the last 20 years, very few have effectively and thoroughly addressed the issue that is common in all school shootings: easy access to guns by those at risk of committing harm.  It has been found that two thirds of school shooters obtained their firearms from their own home, or the home of a relative $^{[8]}$. Studies have shown that the firearm homicide and suicide rates in the US are several times higher than that of any other industrialized country $^{[9]}$, and the patterns appear to be due to higher rates of firearm ownership in the US compared with other industrialized countries $^{[10]}$. It was shocking to learn that the pediatric firearm mortality rate is five times higher in the United States compared with any other industrialized country$^{[11]}$, and 87% of all children age 0 to 14 worldwide killed by firearms are children living in the US $^{[9]}$, despite the fact that less than 5% of the world’s children live in the US $^{[12]}$. Studies have confirmed that there has been unprecedented proliferation of gun rights and gun crimes in the United States in current years and many gun victims are exposed to violence in ways that are accidental, incidental, relational, or environmental $^{[13]}$ which questions the effectiveness of the exisiting gun  policies in different States in the United States. 



## Hypothesis Questions
School shooting is a complex phenomen as it is connected with various crucial components such as: mental health, bullying, school saftey measures, socio-economic condition and most importantly availibilty of guns. The goal of the project was to investigate the impact of gun policies on the school schooting incidences for different states. The analysis was designed to provide valuable insights to the state legislators about the impact of gun policies on the school shooting incidents. In order to analyze school shooting in depth, we scrutinized the following aspects of School Shooting within each state over the time period: 

1. Overall Impacts of the gun laws\

2. Affects of bullying laws and prevalence of bullying\

3. Prevalence of mental illness\

4. Prevalence of  Bullying\

5. School Safety Measures\

6. Affects of Media covergare\

7. Impact of hostility of school environment\

8. Impact of poverty prevalence rate\

9. Impact of specific gun laws\




## References
1. Walker C. (2019, July) 10 years. 180 school shootings. 356 victims. Retrieved from: https://www.cnn.com/interactive/2019/07/us/ten-years-of-school-shootings-trnd/ 

2. Federal Bureau of Investigation. A Study of Active Shooter Incidents in the United States Between 2000 and 2013. 2014; Available at www.fbi.gov/news/stories/2014/september (accessed January 30, 2015).

3. Knowles H (2019, November 15) 16-year-old shooter at California high school has died, authorities say. Retrieved from:https://www.washingtonpost.com/nation/2019/11/15/year-old-santa-clarita-high-school-shooting-suspect-has-died-authorities-say/

4. Towers, S., Gomez-Lievano, A., Khan, M., Mubayi, A., & Castillo-Chavez, C. (2015). Contagion in mass killings and school shootings. PLoS one, 10(7), e0117259.

5. John Woodrow Cox, Steven Rich, Allyson Chiu, John Muyskens and Monica Ulmanu (Updated: 2019, November 14). More than 233,000 students have experienced gun violence at school since Columbine. Retrieved from: https://www.washingtonpost.com/graphics/2018/local/school-shootings-database/

6. Breslau, N., Kessler, R. C., Chilcoat, H. D., Schultz, L. R., Davis, G. C., & Andreski, P. (1998). Trauma and posttraumatic stress disorder in the community: the 1996 Detroit Area Survey of Trauma. Archives of general psychiatry, 55(7), 626-632.

7. Lowe, S. R., Blachman-Forshay, J., & Koenen, K. C. (2015). Epidemiology of trauma and trauma-related disorders: trauma as a public health issue. Evidence-based treatments for trauma-related psychological disorders: A practical guide for clinicians, 11-40.

8. US Dept of Education. Final report and findings of the safe school initiative: Implications for the prevention of school attacks in the United States, 2004.

9. Richardson EG, Hemenway D. Homicide, suicide, and unintentional firearm fatality: comparing the United States with other high-income countries, 2003. The Journal of Trauma and Acute Care Surgery. 2011; 70: 238–243.

10. Hemenway D, Miller M. Firearm availability and homicide rates across 26 high-income countries. Journal of Trauma and Acute Care Surgery. 2000; 49(6): 985–988.

11. Krug, E. G., Mercy, J. A., Dahlberg, L. L., & Powell, K. E. (1998). Firearm-and non-firearm-related homicide among children: an international comparison. Homicide Studies, 2(1), 83-95. 

12. CIA Factbook. The World Factbook; 2010. See also: http://www.ciagov/library/publications/the-world-factbook, accessed January 30, 2015.

13. Metzl, J. M., & MacLeish, K. T. (2015). Mental illness, mass shootings, and the politics of American firearms. American journal of public health, 105(2), 240-249.





